\documentclass{report}
%\addtokomafont{disposition}{\rmfamily}
\usepackage[T1]{fontenc}
\usepackage[a4paper]{geometry}
\usepackage{comment}
\usepackage{afterpage}
\usepackage{xcolor}
\usepackage{wrapfig}
\usepackage[T1]{fontenc}
\usepackage[scaled=.8]{beramono}
\usepackage{booktabs}
\usepackage{tcolorbox}
\usepackage{fullpage}
\usepackage{minted}
\usepackage{pdflscape}
\usepackage[doublespacing]{setspace}
\usepackage{enotez}
\usepackage{subcaption}
%\usepackage[singlelinecheck=off]{caption}
\usepackage{url}
\usepackage{verbatimbox}
\usepackage{listings}
\usepackage{todonotes}
\usepackage[raggedright]{titlesec}
\titlelabel{\thetitle.\quad} % dot after section number
\usepackage{array,multirow,graphicx}
\usepackage{multicol}
\usepackage[hidelinks]{hyperref}
\usepackage{tocloft} 
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage[section]{placeins}
\makeatletter
\AtBeginDocument{%
\expandafter\renewcommand\expandafter\section\expandafter{%
\expandafter\@fb@secFB\section
}%
}
\makeatother
\usepackage[notocbib]{apacite}
\setcounter{tocdepth}{4}
\usepackage{tabularx}
\renewcommand\thesection{\arabic{section}}
\raggedbottom
% column types
\newcolumntype{L}[1]{>{\hsize=#1\hsize\raggedright\arraybackslash}X}%
\newcolumntype{R}[1]{>{\hsize=#1\hsize\raggedleft\arraybackslash}X}%
\newcolumntype{C}[2]{>{\hsize=#1\hsize\columncolor{#2}\centering\arraybackslash}X}%
% reference formatting
\renewcommand{\APACrefatitle}[2]{#1}
\renewcommand{\APACrefbtitle}[2]{#1}
\renewcommand{\APACrefbtitle}[2]{\Bem{#1}}
\renewcommand{\APACrefaetitle}[2]{[#2]}
\renewcommand{\APACrefbetitle}[2]{[#2]}
% possessive cite
\newcommand{\citepos}[1]{\citeauthor{#1}'s \citeyear{#1}}

%\title{\emph{I am not my bipolar}: a corpus-driven discourse analysis of an online support group}
\title{Discourse-semantics of risk in \emph{The New York Times}, 1963, 1987--2014: a corpus linguistic approach}
\author{
Jens O. Zinn\\
\texttt{jzinn@unimelb.edu.au}
\and
Daniel McDonald\\
\texttt{mcdonaldd@unimelb.edu.au}\\
}

\date{University of Melbourne, Australia\\
~\\
\today}

\begin{document}

% words that need line breaks
\hyphenation{under-lying language forums socia-bility dis-cur-sive pred-icator conc-ord-ancing Lee-uwen object-ivity lexico-grammar lexico-gramm-atical meta-function Esco-bedo}

% abstract heading
\renewcommand{\abstractname}{Abstract}

% make title page
\maketitle

\begin{abstract}
\noindent Since the 1980s and 1990s the notion of risk has become increasingly influential in societal discourses and scholarly debate \cite{skolbekken_risk_1995}. From early work on risk and culture \cite{douglas_risk_1986,douglas_risk_2013} to the \emph{risk society} thesis \cite{beck_risk_1992,beck_world_2009,giddens_runaway_2002}, from governmentality theorists working in the tradition of Foucault \cite{dean_governmentality:_2010,omalley_risk_2012,rose_powers_1999} to modern systems theory \cite{luhmann_risk:_1993} all have built their work around the notion of risk and implicitly or explicitly refer to linguistic changes. Though this body of literature offers different explanations for the shift towards \emph{risk} and its connection to social change, to date there has been no attempt to empirically examine their relative ability to explain this change in the communication of possible harm. 

% We will address this deficit by examining a number of claims made by different sociological risk approaches in more detail. We will utilise a corpus based approach to examine in detail how the institutional and sociocultural shift towards risk has manifested linguistically.

To address this defecit, we conduct a corpus-based investigation of risk words in \emph{The New York Times} between 1963 and mid-2014. The investigation involves the creation of an annotated corpus of over 150,000 unqique risk tokens and their co-text. Purpose-built functions for manipulating this dataset and visualising results were created and used to investigate the corpus according to a systemic-functional conceptualisation of the transitivity and mood systems of language. This toolkit is freely available at \url{https://github.com/interrogator/corpkit}. Following the corpus interrogation, we use functional linguistics and sociological risk theory in tandem to analyse the findings. First, systemic-functional linguistics is used to link lexicogrammatical phenomena to discourse-semantic meaning of the texts. Longitudinal changes in risk language are then mapped to key events, as well as broader social movements.

This report is accompanied by an interactive \emph{IPython Notebook} interface to our corpus and developed computational tools. Key findings from this report are stored there, as well as additional information (e.g. concordance lines, keywords, collocations), that could not be included in this report due to spatial considerations. It is available for both interactive and static viewing at \url{https://github.com/interrogator/risk}.

\end{abstract}
\cleardoublepage
\pagenumbering{roman}
%\setlength{\cftbeforetoctitleskip}{-3em}
\singlespacing
\tableofcontents
\cleardoublepage
\listoftables
\cleardoublepage
\listoffigures
\cleardoublepage
\onehalfspacing
\pagenumbering{arabic}

% summary without toc entries
\section*{Executive summary}

\noindent Since the middle of the 1980s sociologists have started to argue that after WW2 significant social transformations took place in most Western industrialised societies which have manifested in a shift towards the risk semantic. Risk has become pervasive in public and scholarly debates and practices in Europe and elsewhere in the world. Following the common assumption that societal changes and language changes are closely linked, this research report aims to contribute to advancing understanding of this social shift towards risk. 

There is a wealth of literature and several sociological theories (e.g. risk society, socio-cultural, governmentality, systems theory, edgework) which offer different explanations for the shift towards risk and its connection to social change. Yet, to date there has been little attempts to empirically examine their relative ability to explain this change in the communication of possible harm. This research report addresses this deficit by:
%
\begin{enumerate}
\item Examining a number of claims made by different sociological theories and exploring linguistic changes which might not have been theoretically addressed yet. 
\item Developing a corpus based research strategy and computational research tools to examine in detail how the institutional and sociocultural shift towards risk has manifested linguistically.
\end{enumerate}
%
The study utilises The New York Times (NYT) as a case study for the US to reconstruct the growing usage of the term risk from 1987 to 2014 and examines how discourse-semantic shifts are linked to institutional and socio-cultural changes as well as socially relevant events (e.g. crises and disasters). In addition the study uses a sample of volume 1963 articles of the NYT to contrast the results with much earlier ways of risk reporting.

The investigation involves the creation of an annotated corpus of over 150,000 unique risk tokens and their co-text. Purpose-built functions for manipulating this dataset and visualising results were created and used to investigate the corpus according to a systemic functional conceptualisation of the transitivity and mood systems of language. This toolkit is freely available at \url{https://github.com/interrogator/corpkit}. 

Following the corpus interrogation, we use functional linguistics and sociological risk theory in tandem to analyse the findings. First, systemic functional linguistics is used to link lexicogrammatical phenomena to discourse-semantic meaning of the texts. Longitudinal changes in risk language are then mapped to key events, as well as broader social processes and changes.

This report is accompanied by an interactive IPython Notebook interface to our corpus and developed computational tools. Key findings from this report are stored there, as well as additional information (e.g. concordance lines, keywords, collocations), that could not be included in this report due to spatial considerations. It is available for both interactive and static viewing at \url{https://github.com/interrogator/risk}.

\section*{Research question, hypotheses} 

Most generally, the study examines \emph{how the institutionalisation of new societal practices manifest linguistically in the change of risk discourses and the use of risk language}. We also examine a number of hypotheses derived from mainstream sociological risk theorising:

\begin{enumerate}
\item Is there a shift in discourses from the positive notion of risk-taking to the negative meaning of risk?
\item Can we observe an increasing salience of the at-risk status of social groups?
\item Is there evidence for the assumption of individualised experience of risk?
\item Is there any evidence for the decreasing calculability/controllability of risk?
\item Are there any differences between different social groups in the exposure to risk?
\end{enumerate}

\section*{Outcomes}

\begin{enumerate}
\item The linguistic analyses on the lexicogrammatical level and in discourse semantics give clear indications for a growing routinisation and institutionalisation of risk. We found a growing number of forms such as risk assessment or risk regulator which indicate the institutionalisation of risk practices. Decreasing arguablity and growing implicitness of risk in the clause indicated a shift from the centre to the ancillary parts of the sentence.
\item There are also clear hints that the negative notion of risk is becoming more common, indicated by a clear tendency to nominalisation but also to processes where agency is minimal (e.g. put at-risk).
\item There is a clear trend towards decreasing agency in risk processes from risking and taking risk to running risk and putting at-risk. Thus the strong norm of individual responsibility and risk-taking is accompanied by media-coverage which emphasises lacking agency in risk.
\item There is a clear increase of reporting on issues where people and particular social groups are presented as lacking control especially regarding health issues.
\item Risk reporting, in particular in the health sector, is driven by reference to scientific research supporting a rationalised approach to risk with formalised concepts such as risk factors and reference to research terms on the rise. However, expressions of control such as calculated risk are decreasing while expressions indicating the possibility of negative outcomes are increasing indicating a rise of a possibilistic notion of risk.
\item New coverage in the NYT shows a clear difference between powerful risk-takers and relatively powerless at-risk groups. The difference between the groups is sharpening over time. The powerful take more social risks while the powerless take much more substantial risks often related to illness, injury and death.
\end{enumerate}

\section*{Key methodological issues}

The study demonstrates the potential for large annotated corpora to examine key claims of sociological theory about social change. It is innovative in creating a corpus comprised by annual subcorpora which allows detailed longitudinal interrogations. It takes advantage of major developments in computational linguistics\slash natural language processing such as constituency and dependency parsing. Both allow much more detailed and nuanced investigations than generally seen in CADS.

The advantages of the longitudinal analysis of one newspaper had a price, however. It is limited to a case of one genre of newscoverage. Other forms of text are not considered and might show different patterns. 

The Language processing tools developed and\slash or used in this study also have limitations. We have used the \emph{Stanford CoreNLP} NLP suite, which outputs constituency and dependency parses, rather than systemic-functional parsers. Translating the former into the latter poses a number of serious challenges. In some cases, systemic features cannot be automatically recovered from the parses, limiting the kinds of phenomena that can be automatically located and\slash or counted.

% Automated querying of lexicogrammatical patterns in of large data sets have limits due to the complexity of linguistic expressions of meaning. It necessarily has to focus on the more common patterns.

\section*{Perspectives} % not sure about this heading title

At the time of writing, our investigation is ongoing. In forthcoming work, we seek to:

\begin{enumerate}
\item Broaden the empirical basis integrating a growing number of US newspapers, in order to determine whether findings generated here may be generalised across mainstream U.S. print media.
\item Integrate data from newspapers of other countries (e.g. UK, Australia) into our corpus, in order to identify differences in institutional and linguistic practices across Anglophone countries.
\item Perform qualitative analysis of indivudal texts, in order to explore particular observations made in this report and their institutional contexts in more detail.
\end{enumerate}

\include{chapters/chapter_01} % Introduction

\include{chapters/chapter_02} % Case study

\include{chapters/chapter_03} % Methodology

\include{chapters/chapter_04} % Findings: annual

\include{chapters/chapter_05} % Findings: health corpus

\include{chapters/chapter_06} % Discourse semantics in NYT

\include{chapters/chapter_07} % Limitations, research agenda, conclusions

%end matter

\cleardoublepage
\singlespacing
%\footnotesize
\printendnotes
\cleardoublepage
\singlespacing
\bibliographystyle{apacite}
\bibliography{references/references}
\cleardoublepage

\end{document}
